Parametricity Consequences of GADTs Representations

### 2021-02-20 20-gadts/GADT-story.tex
work on gadt-story

GADTs are traditionally represented via Church encodings. In this
view, GADTs are completely determined by their syntax; for example,
the GADT $\mathsf{G}$ defined in Haskell by
\begin{equation}\label{eq:example-GADT}
\begin{array}{l}
\mathsf{data\, Expr\,a\,where}\\
\mathsf{\;\;\;\;\;\;\;\;Const ::\, a \to Expr\,a}\\
\mathsf{\;\;\;\;\;\;\;\;Pair\,\,\,\,\, ::\, Expr \,a \to Expr\,b \to
  Expr\,(a \times b)}\\ 
\end{array}
\end{equation}
%\[\begin{array}{l}
%\mathsf{data\,G : * \to *\,where}\\
%\mathsf{\;\;\;\;\;\;\;\;c : G\,1}
%\end{array}\]
%{\color{blue} Use $\top$ instead of $1$ throughout?}
\noindent
comprises expressions of any type $\mathsf{a}$ and expressions
obtained by pairing other expressions. By contrast, the primitive
representation sees a GADT as the {\em functorial
  completion}~\cite{jp19} of its syntax. Completing a GADT's syntax to
a {\em functor} is necessary in order to be able to take the least
fixed point~\cite{tfca}, so it is inherent in the view of data types
as their primitive representations.  Viewed as its primitive
representation, $\mathsf{Expr}$ comprises not just those data elements
representable in syntax, but also all data elements of the form
$\mathsf{map\,f\,e}$ for all types $\mathsf{t_1}$ and $\mathsf{t_1}$,
all functions $\mathsf{f : t_1 \to t_2}$ definable in the language,
and all $\mathsf{e :: Expr\,t_1}$, as well as all data elements of the
form $\mathsf{map\,g\,e'}$ for each appropriately typed function
$\mathsf{g}$ and each element $\mathsf{e'}$ already added to the data
type, and so on. Functorial completion adds no new data elements to
the syntax of a GADT that is actually a (polynomial) ADT or a
(hereditarily polynomial) nested type, so its Church encoding and
primitive representation coincide. However, these representations are
rarely the same for GADTs.
%Indeed, they are quite different for even the very simple GADT
%$\mathsf{G}$ above.

@@

{\color{red} This will probably be more convincing if we use, e.g.,
  an expression GADT from Stephanie above.}
A GADT programmer is likely to use GADTs precisely because they
exhibit different behaviors at different types, and thus to consider
the GADT to be completely specified by its syntax. When used this way,
the shape of a GADTs is actually {\em determined} by the data it
contains. As a result, it is not expected to support a $\mathsf{map}$
function --- i.e., to be an instance of Haskell's $\mathsf{Functor}$
class --- like other data types. For example, the definition for
$\mathsf{G}$ above specifies data only at instance $\mathsf{G\,1}$.
Since there is a function from $\mathsf{1}$ to any other type
$\mathsf{t}$, and yet $\mathsf{G\,t}$ is uninhabited at any instance
other than $\mathsf{1}$, $\mathsf{G}$ cannot support a $\mathsf{map}$
function. When it is intended to capture this kind of non-uniform
behavior, a GADT's behavior does not generalize that of ADTs as
advertised, and in some sense doesn't even behave like a traditional
data type.  When used this way, a GADT cannot be seen as the least
fixpoint of a functor, and must instead be represented via its Church
encoding.

@@

Importantly, $\mathit{Equal}\,a$ cannot be made into a functor, or,
equivalently, $\mathsf{Equal\,a}$ cannot be made an instance of
Haskell's $\mathsf{Functor}$ class: if $\mathsf{Equal\,a}$ supported a
function
\[\mathsf{map_{Equal\,a} :: (b \to c) \to Equal\,a\,b \to
  Equal\,a\,c}\]
then defining
\[\begin{array}{l}
\mathsf{eqElim :: Equal\, a\, b \to b \to a}\\
\mathsf{eqElim\, Refl\, x = x} 
\end{array}\]
would allow us to construct the element 
\[\begin{array}{l}
\mathsf{emptyFromMapEquala :: (\forall b\, c.\, (b \to c) \to Equal\, a\,
  b \to Equal\, a \,c) \to Void}\\ 
\mathsf{emptyFromMapEquala \,map_{Equal\,a} = eqElim\, (map_{Equal\,a}\,
  absurd\, Refl)\, ()}  
\end{array}\]
of type $\mathsf{Void}$, which is impossible. It follows immediately
from this contradiction that $\mathsf{Seq}$ cannot support a map
function either. This is already intuitively evident from the Haskell
definition of $\mathsf{Seq}$, though. Indeed, the clause of
$\mathsf{map}$ for the $\mathsf{Pair}$ constructor should feed
$\mathsf{map}$ a function $\mathsf{f :: (b \times c) \to e}$ and a
term of the form $\mathsf{Pair \,t_1\,t_2}$ for $\mathsf{x :: b}$ and
$\mathsf{y :: c}$, and produce a term $\mathsf{Pair\,t_1'\,t_2'}$ for
some appropriately typed terms $\mathsf{t_1'}$ and
$\mathsf{t_2'}$. But it is not clear how to achieve this since
$\mathsf{e}$ need not necessarily be a product type. And even if
$\mathsf{e}$ {\em were} known to be of the form $\mathsf{w \times z}$,
then we still wouldn't necessarily have a way to produce data of type
$\mathsf{w \times z}$ from only $\mathsf{f :: b \times c \to w \times
  z}$ and $\mathsf{x}$ and $\mathsf{y}$ unless we knew, e.g., that
$\mathsf{f}$ was a pair of functions.

>>>>

GADTs are traditionally represented via Church encodings. In this
view, GADTs are completely determined by their syntax; for example,
the GADT $\mathsf{G}$ defined in Haskell by
\begin{equation}\label{eq:example-GADT}
\begin{array}{l}
\mathsf{data\, Seq\,a\,where}\\
\mathsf{\;\;\;\;\;\;\;\;Const ::\, a \to Seq\,a}\\
\mathsf{\;\;\;\;\;\;\;\;Pair\,\,\,\,\, ::\, Seq \,a \to Seq\,b \to
  Seq\,(a \times b)}\\ 
\end{array}
\end{equation}
%\[\begin{array}{l}
%\mathsf{data\,G : * \to *\,where}\\
%\mathsf{\;\;\;\;\;\;\;\;c : G\,1}
%\end{array}\]
%{\color{blue} Use $\top$ instead of $1$ throughout?}
\noindent
comprises sequences of any type $\mathsf{a}$ and sequences obtained by
pairing the data in two already existing sequences. By contrast, the
primitive representation sees a GADT as the {\em functorial
  completion}~\cite{jp19} of its syntax. Completing a GADT's syntax to
a {\em functor} is necessary in order to be able to take the least
fixed point~\cite{tfca}, so it is inherent in the view of data types
as their primitive representations. Viewed as its primitive
representation, $\mathsf{Seq}$ comprises not just those data elements
representable in syntax, but also all data elements of the form
$\mathsf{map\,f\,s}$ for all types $\mathsf{t_1}$ and $\mathsf{t_2}$,
all functions $\mathsf{f : t_1 \to t_2}$ definable in the language,
and all $\mathsf{s :: Seq\,t_1}$, as well as all data elements of the
form $\mathsf{map\,g\,s'}$ for each appropriately typed function
$\mathsf{g}$ and each element $\mathsf{s'}$ already added to the data
type, and so on. Functorial completion for $\mathsf{Seq}$ adds, in
particular, data elements of the form
$\mathsf{map\,g\,(Pair\,u_1\,u_2)}$ even though these may not
themselves be of the form $\mathsf{Pair\,v_1\,v_2}$ for any terms
$\mathsf{v_1}$ and $\mathsf{v_2}$. By contrast, functorial completion
adds no new data elements to the syntax of a GADT that is actually a
(polynomial) ADT or a (hereditarily polynomial) nested type, so its
Church encoding and primitive representation coincide. However, these
representations are rarely the same for GADTs.

@@

A GADT programmer is likely to use GADTs precisely because they
exhibit different behaviors at different types, and thus to consider a
GADT to be completely specified by its syntax. When used this way, the
shape of a particular element of a GADTs is actually {\em determined}
by the data it contains. As a result, it is not expected to support a
$\mathsf{map}$ function --- i.e., to be an instance of Haskell's
$\mathsf{Functor}$ class --- like other data types. Indeed, the
definition for $\mathsf{Seq}$ above specifies that an element of the
form $\mathsf{Pair\,u_1\,u_2}$ must have the shape of a sequence of
data of pair type rather than a sequence of data of arbitrary type
$\mathsf{e}$.  The clause of $\mathsf{map}$ for the $\mathsf{Pair}$
constructor should therefore feed $\mathsf{map}$ a function $\mathsf{f
  :: (a \times b) \to e}$ and a term of the form $\mathsf{Pair
  \,u_1\,u_2}$ for $\mathsf{u_1 :: Seq\,a}$ and $\mathsf{_2 ::
  Seq\,b}$, and produce a term $\mathsf{Pair\,v_1\,v_2'}$ for some
appropriately typed terms $\mathsf{v_1}$ and $\mathsf{v_2}$. However,
it is not clear how to achieve this since $\mathsf{e}$ need not
necessarily be a product type. {\color{blue} And even if $\mathsf{e}$
  {\em were} known to be of the form $\mathsf{w \times z}$, then we
  still wouldn't necessarily have a way to produce data of type
  $\mathsf{w \times z}$ from only $\mathsf{f :: a \times b \to w
    \times z}$ and $\mathsf{u_1}$ and $\mathsf{u_2}$ unless we knew,
  e.g., that $\mathsf{f}$ was a pair of functions.}

{\color{red} HERE!}

data only at instance
$\mathsf{G\,1}$.  Since there is a function from $\mathsf{1}$ to any
other type $\mathsf{t}$, and yet $\mathsf{G\,t}$ is uninhabited at any
instance other than $\mathsf{1}$, $\mathsf{G}$ cannot support a
$\mathsf{map}$ function. When it is intended to capture this kind of
non-uniform behavior, a GADT's behavior does not generalize that of
ADTs as advertised, and in some sense doesn't even behave like a
traditional data type.  When used this way, a GADT cannot be seen as
the least fixpoint of a functor, and must instead be represented via
its Church encoding.

@@

Importantly, $\mathit{Equal}\,a$ cannot be made into a functor, or,
equivalently, $\mathsf{Equal\,a}$ cannot be made an instance of
Haskell's $\mathsf{Functor}$ class: if $\mathsf{Equal\,a}$ supported a
function
\[\mathsf{map_{Equal\,a} :: (b \to c) \to Equal\,a\,b \to
  Equal\,a\,c}\]
then defining
\[\begin{array}{l}
\mathsf{eqElim :: Equal\, a\, b \to b \to a}\\
\mathsf{eqElim\, Refl\, x = x} 
\end{array}\]
would allow us to construct the element 
\[\begin{array}{l}
\mathsf{emptyFromMapEquala :: (\forall b\, c.\, (b \to c) \to Equal\, a\,
  b \to Equal\, a \,c) \to Void}\\ 
\mathsf{emptyFromMapEquala \,map_{Equal\,a} = eqElim\, (map_{Equal\,a}\,
  absurd\, Refl)\, ()}  
\end{array}\]
of type $\mathsf{Void}$, which is impossible. It follows immediately
from this contradiction that $\mathsf{Seq}$ cannot support a map
function either. This is already intuitively evident from the Haskell
definition of $\mathsf{Seq}$, though. ALREADY USED??? Indeed, the clause of
$\mathsf{map}$ for the $\mathsf{Pair}$ constructor should feed
$\mathsf{map}$ a function $\mathsf{f :: (b \times c) \to e}$ and a
term of the form $\mathsf{Pair \,t_1\,t_2}$ for $\mathsf{x :: b}$ and
$\mathsf{y :: c}$, and produce a term $\mathsf{Pair\,t_1'\,t_2'}$ for
some appropriately typed terms $\mathsf{t_1'}$ and
$\mathsf{t_2'}$. But it is not clear how to achieve this since
$\mathsf{e}$ need not necessarily be a product type. And even if
$\mathsf{e}$ {\em were} known to be of the form $\mathsf{w \times z}$,
then we still wouldn't necessarily have a way to produce data of type
$\mathsf{w \times z}$ from only $\mathsf{f :: b \times c \to w \times
  z}$ and $\mathsf{x}$ and $\mathsf{y}$ unless we knew, e.g., that
$\mathsf{f}$ was a pair of functions.
